### 2017-07-07 paper/compositional.tex
small changes

Hence, NCSMs effectively trade the uniform integrability condition for 
non-negativity over the whole program. We believe that the latter condition is 
substantially easier to impose and check automatically, especially in the case 
of affine probabilistic programs and (propositionally) linear NCSMs: for 
\APP{}s, synthesizing NCSM entails synthesizing sufficient program invariants 
(for which there is a good automated tool support~\cite{FG10:aspic})  encoding 
the ranking, 
unaffection, and non-negativity conditions into a collection of linear 
constraints (as for general LinLexRSMs in Section~\ref{sec:algo}). 
Figures~\ref{fig:uniint} and~\ref{fig:uniint2} show instances where attempts to 
to prove of a.s. termination via PVSMs fail while proofs via LinNCSMs work.

\begin{example}
For all programs in Figures~\ref{fig:uniint} and~\ref{fig:uniint2} there are 
LinNCSMs for all the loops in the program, which shows that the programs 
terminate 
a.s. In Figure~\ref{fig:uniint}, for both programs the inner loops have 
LinNCSMs of the form $c+d_{\loc}$, for $d_{\loc}$ a location-specific constant, 
while the outer loops have LinNCSMs of the form $x+d_{\loc}'$. In 
Figure~\ref{fig:uniint2} the program similarly has LinNCSMs defined, proceeding 
from the innermost loop and neglecting the location-specific constants, by 
variables $z$, $y$, $x$.
\end{example}
%While we defined NCSMs as one-dimensional measurable maps to make them analogous to

@@

\Input{An \APP{} $\program$ together with an invariant map $\inv$.}
$d\leftarrow \text{ depth of loop nesting in $\program$}$\;
\For{$i \leftarrow d$ \KwTo $0$}{
\ForEach{sub-loop $\program'$ of $\program$ nested $i$ levels below the main loop}{
\label{algoline:one}
$\linsystem \leftarrow$ system of lin. constraints encoding the existence of LinNCSM for $\program'$ supported by $\inv$\;
\label{algoline:two}\If{$\linsystem$ not solvable}{check existence of a PVSM for $\program'$~\cite{HolgerPOPL}
\;
\lIf{PVSM does not exist}{\Return{``cannot prove a.s. termination of $\program$''}}
}
}
}
\Return{``$\program$ terminates a.s.'' }
\caption{Compositional Termination Proving}
\label{algo:compositional}
\end{algorithm}

>>>>

%Hence, NCSMs effectively trade the uniform integrability condition for 
%non-negativity over the whole program. We believe that the latter condition is 
%substantially easier to impose and check automatically, especially in the case 
%of affine probabilistic programs and (propositionally) linear NCSMs: 
For 
\APP{}s, synthesizing linear NCSM entails synthesizing sufficient program 
invariants 
(for which there is a good automated tool support~\cite{FG10:aspic}) and  
encoding 
the ranking, 
unaffection, and non-negativity conditions into a collection of linear 
constraints (as for general LinLexRSMs in Section~\ref{sec:algo}). 
Figures~\ref{fig:uniint} and~\ref{fig:uniint2} show instances where attempts to 
to prove of a.s. termination via PVSMs fail while proofs via LinNCSMs work.

\begin{example}
All programs in Figures~\ref{fig:uniint} and~\ref{fig:uniint2} have 
LinNCSMs for all their loops, which shows that the programs 
terminate 
a.s. In Figure~\ref{fig:uniint}, for both programs the inner loops have 
LinNCSMs of the form $c+d_{\loc}$, for $d_{\loc}$ a location-specific constant, 
while the outer loops have LinNCSMs of the form $x+d_{\loc}'$. In 
Figure~\ref{fig:uniint2} the program similarly has LinNCSMs defined, proceeding 
from the innermost loop and neglecting the location-specific constants, by 
variables $z$, $y$, $x$.
\end{example}
%While we defined NCSMs as one-dimensional measurable maps to make them analogous to

@@

\Input{An \APP{} $\program$ together with an invariant map $\inv$.}
%$d\leftarrow \text{ depth of loop nesting in $\program$}$\;
\For{$i \leftarrow \text{ depth of loop nesting in $\program$}$ \KwTo $0$}{
\ForEach{sub-loop $\program'$ of $\program$ nested $i$ levels below the main loop}{
\label{algoline:one}
$\linsystem \leftarrow$ system of lin. constraints encoding the existence of LinNCSM for $\program'$ supported by $\inv$\;
\label{algoline:two}\If{$\linsystem$ not solvable}{check existence of a PVSM for $\program'$~\cite{HolgerPOPL}
\;
\lIf{PVSM does not exist}{\Return{``cannot prove a.s. termination of $\program$''}}
}
}
}
\Return{``$\program$ terminates a.s.'' }
\caption{Compositional Termination Proving}
\label{algo:compositional}
\end{algorithm}

### 2017-07-07 paper/extensions.tex
small changes

As shown in Example~\ref{ex:infinite-time}, already LinLexRSM maps are capable 
of proving almost-sure termination of programs whose expected termination time 
is infinite. However, it is often desirable to obtain bounds on expected 
runtime of a program. In this section, we present a LexRSM-based method for 
obtaining bounds on expected runtime, and we show how to automatize the usage 
of this method to obtain bounds on expected runtime in a 
subclass of \PP{}s.

As in the case of a.s. termination we start with general mathematical statement 
about LexRSMs. We define a restricted class of strict LexRSMs with \emph{bounded 
expected conditional increase} property. Recall from 
definition~\ref{def:lexrsm} that strict
LexRSM for a stopping time $\stime$ is characterized by the possibility to 
a.s. partition, for each $i\in \Nset_0$, the set $\{\omega\in \Omega\mid 
\stime(\omega)>i \}$ into $n$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
on $L^i_j$ the conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is smaller than $\vec{X}_i[j]$, and for all $j'<j$, on $L^i_{j'}$ the 
conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is no larger than $\vec{X}_i[j]$. This leaves the opportunity of conditional 
expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
$L^i_{j''}$ with $j''>j$. The conditional expected increase property bounds the 
possibility of this increase.

@@

>>>>

As shown in Example~\ref{ex:infinite-time}, already LinLexRSM maps are capable 
of proving almost-sure termination of programs whose expected termination time 
is infinite. However, it is often desirable to obtain bounds on expected 
runtime of a program. In this section, we present a LexRSM-based method for 
obtaining bounds on expected runtime using a restricted class of LexRSMs.

As in the case of a.s. termination we start with general mathematical statement 
about LexRSMs. We define a restricted class of strict LexRSMs with \emph{bounded 
expected conditional increase} property. Recall from 
definition~\ref{def:lexrsm} that strict
LexRSM for a stopping time $\stime$ is characterized by the possibility to 
a.s. partition, for each $i\in \Nset_0$, the set $\{\omega\in \Omega\mid 
\stime(\omega)>i \}$ into $n$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
on $L^i_j$ the conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is smaller than $\vec{X}_i[j]$, and for all $j'<j$, on $L^i_{j'}$ the 
conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is no larger than $\vec{X}_i[j]$. This leaves the opportunity of conditional 
expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
$L^i_{j''}$ with $j''>j$. The expected conditional increase (ECI) property 
bounds the 
possibility of this increase.

@@

Previous work testifies that analogous ``bounded 
expected change'' constraints naturally emerge when one aims to obtain 
additional information about runtime of programs 
~\cite{HolgerPOPL,CFNH16:prob-termination,CNZ17,CF17}. The major novelty of our 
bounded ECI condition is that it can be used to obtain polynomial bounds on 
expected runtime using \emph{linear} supermartingales.

### 2017-07-07 paper/prelims.tex
small changes

\vspace{-1em}
\subsection{Basic Notions}
\vspace{-0.5em}
For a set $A$ we denote by $|A|$ the cardinality of $A$. We denote by $\Nset$,
$\Nset_0$, $\Zset$, and $\Rset$ the sets of all positive integers, non-negative
integers, integers, and real numbers, respectively. We assume basic knowledge 
of matrix calculus.
We use boldface notation for
vectors, e.g. $\vec{x}$, $\vec{y}$, etc., and we denote an $i$-th component of a
vector $\vec{x}$ by $\vec{x}[i]$. 
%For the purpose of matrix calculations we 
%assume that (non-transposed) vectors are row vectors. If $\vec{v},\vec{v}'$ 
%are 
%$n$ and $m$ dimensional vectors, respectively, then $(\vec{v},\vec{v}')$ is an 
%$(n+m)$-dimensional vector obtained by ``concatenation'' of $\vec{v}$ and 
%$\vec{v}'$. 
We identify 1-dimensional vectors with numbers. For an 
$n$-dimensional vector 
$\vec{x}$, index $1 \leq i\leq n$, and number $a$ we denote by $\vec{x}(i\leftarrow a)$ 
a 
vector $\vec{y}$ such that $\vec{y}[i]=a$ and $\vec{y}[j]=\vec{x}[j]$ for all 
$1\leq j \leq n$, $j\neq i$.
For comparison of vectors (e.g. as in $\vec{x}\leq \vec{y}$), we consider componentwise 
comparison. 
For comparing functions $f,g$ with the same domains, we write $f\leq g$ 
if $f(x)\leq g(x)$ for all $x$ in the domain.

\smallskip\noindent{\em Variables.}
Throughout the paper we fix a countable set of variables $\vars$. 
We consider some arbitrary but fixed linear order on the set of all variables, 
hence we write 
$\vars=\{x_1,x_2,x_3,\dots\}$. 
%A \emph{valuation} is a partial function 
%$\val\colon 
%\vars \dashrightarrow \Rset$ assigning a unique number to some of the 
%variables. A \emph{domain} of a valuation $\val$, denoted by %$\dom(\val)$, is 
%the set of all variables $x$ such that $\val(x)$ is defined. Given a %set 
%$\pvars\subseteq \vars$ of variables, we denote by $\vals(\pvars)$ the %set of 
%all valuations $\val$ with $\dom(\val)=\pvars$.

@@

\smallskip\noindent{\em Expressions.}
An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
is an expression in the standard programming-language sense, i.e. a formula 
built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
numerical operators from some fixed finite set. Each expression $E$ over 
$\{x_1,\dots,x_n\}$ 
determines a function which for each $m$-dimensional vector $\vec{x}$, where 
$m\geq n$,  
returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
Slightly abusing our notation, we denote this function also by $E$ and the 
value of this function on argument $\vec{x}$ by $E(\vec{x})$. We do not a priori fix a concrete set of operators that can be used to form expressions. However, in order to ensure that semantics of probabilistic programs with real-valued variables is defined correctly, we impose the following two conditions on the set of expressions used in each program:
%\begin{enumerate}
(1)
For each expression $E$ over variables $\{x_1,\dots,x_n\}$ and each $n$-dimensional vector $\vec{x}$ the value $E(\vec{x})$ is well defined.\footnote{Our results can be easily extended to programs where encountering an expression of undefined value, such as division by zero, triggers an exception which terminates the program, but we abstract away from such details for the sake of clarity.}
(2)
The function defined by each expression $E$ is Borel-measurable (for definition of Borel-measura\-bility, see, e.g.~\cite{Billingsley:book}).
%\end{enumerate}

From measure theory it is known that these conditions hold in particular for programs where expressions are build using the standard arithmetic operators of addition, subtraction, multiplication, and division (provided that expressions evaluating to zero are not allowed as divisors).

@@

\smallskip\noindent{\em Probabilistic Control Flow Graphs.}
%There are several ways in which one can express the semantics of \APP's with
%(angelic and demonic) 
%non-determinism
We consider standard operational semantics of \PP{}s defined via an 
uncountable state-space
Markov decision process (MDP) (uncountable due to real-valued variables).
That is, we associate to each program 
a certain stochastic process.
To define this process, we first define so called 
\emph{probabilistic control flow graphs}~\cite{CFG16}.

\smallskip
\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinitset,\transitions,\updates,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into four 
pairwise
disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
non-deterministic branching, 
probabilistic branching, deterministic, and assignment locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinitset$ is a set of initial 
\emph{assignment vectors};
% POKRACOVAT ODSUD
\item $\transitions\subseteq\locs\times\locs$ is a transition relation;
%whose members are tuples of
%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
%variable index}, and $u$ 
%is an 
%\emph{update element}, which can 
%be one of the following mathematical objects: %\begin{itemize}
%%\item 
%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%%\item 
%(b)~a distribution $d\in \mathcal{D}$; or
%%%\item 
%(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\updates$ is a function assigning to each transition outgoing from an 
assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
\emph{target 
	variable index} and $\up$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~a Borel-measurable function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%\item 
(c)~a set $R\subseteq \Rset$ (representing a non-deterministic update).
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$;
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

We assume that each location has at least one outgoing transition.
Also, for every deterministic location $\ell$ we assume the following: if
$\tau_1,\dots,\tau_k$ are all transitions outgoing from $\ell$, then $G(\tau_1)
\vee \dots \vee G(\tau_k) \equiv \mathit{true}$ and $G(\tau_i) \wedge G(\tau_j)
\equiv \mathit{false}$ for each $1\leq i < j \leq k$. For each 
distribution $d$ appearing in the 
pCFG we assume the following features are known: expected value $\expv[d]$ of 
$d$ and a set $SP_d$ containing the \emph{support} of $d$. The support is the  
smallest 
closed set of real numbers whose complement has probability zero 
under $d$.\footnote{In particular, a support of a \emph{discrete} probability 
	distribution $d$ is simply the at most countable set of all points on a 
	real 
	line that have positive probability under $d$.} Finally, we assume that 
	each 
assignment location has at most (and thus exactly) one outgoing transition. The 
translation from probabilistic programs to the corresponding pCFG is 
standard~\cite{CFNH16:prob-termination}, and the details are presented 
in~\AppendixMaterial.
%Finally, we assume that for each 
%transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
%deterministic. This is just a technical assumption yielding no loss of 
%generality, and it somewhat simplifies notation.
%And we assume that each coordinate of $\mathcal{D}$ represents an integrable
%random variable (i.e., the expected value of the absolute value of the random
%variable exists).
\end{definition}

\smallskip\noindent{\em Configurations.}
A \emph{configuration} of a pCFG $\pCFG$ is a tuple $(\ell,\vec{x})$,
where $\ell$ is a location of $\pCFG$ and $\vec{x}$ is an 
$|\pvars|$-dimensional vector.
We say that a transition $\tau$ is \emph{enabled} in a configuration
$(\ell,\vec{x})$ if $\ell$ is the source location of $\tau$ and in addition,
${\vec{x}}\models G(\tau)$ provided that $\ell$ is deterministic. 
%A 
%configuration 
%$(\loc,\vec{x})$ is non-deterministic/probabilistic/deterministic/assignment 
%one if 
%$\loc$ is 
%non-deterministic/probabilistic/deterministic/assignment one, respectively.

\smallskip\noindent{\em Executions and reachable configurations.}
We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
configuration $(\loc,\vec{x})$ if 
%A \emph{finite path} (or
%\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
%configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
%for each
%$0 \leq i < k$ 
there is a transition  $\tau=(\loc,\loc')$ enabled in
$(\loc,\vec{x})$ 
%such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
%satisfies one of the following:
and $\vec{x}'$ satisfies the following:
\begin{compactitem}
\item if $\loc$ is not an assignment location, then $\vec{x}'=\vec{x}$;
\item if $\loc$ is an assignment location with $\updates(\tau)=(j,\up)$, then 
$\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ satisfies one of the 
following 
depending on the type of $\up$:
%\begin{compactitem}
%\item 
(a)~if $\up$ is a Borel-measurable function, then $a=\up(\vec{x})$;
%\item 
(b)~if $\up$ is an integrable\footnote{A distribution on some numerical 
domain 
is integrable if its expected value exists and is finite. In particular, each 
Dirac distribution is integrable.} distribution $d$, then 
$a\in \support(d)$; and
%\item
(c)~if $\up$ is a set, then $a$ is some element of $\up$.
%\end{compactitem}
\end{compactitem}
A \emph{finite path} (or
\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
$\loc_0=\locinit$, $\vec{x}_0\in\vecinitset$, and 
%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
for each
$0 \leq i < k$ the configuration $(\loc_{i+1},\vec{x}_{i+1})$ is a successor of 
$(\loc_i,\vec{x}_i)$.
A \emph{run} (or \emph{execution}) in
$\pCFG$ is an infinite sequence of configurations whose every finite
prefix is a finite path.
A configuration $(\loc,\vec{x})$ is {\em reachable} from the initial 
configuration
$(\locinit,\vecinit)$ (where, $\vecinit\in \vecinitset$)
if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
$(\loc,\vec{x})$. We denote by $\Conf_\pCFG,\Fpath_\pCFG$ and $\Run_\pCFG$ the 
sets of all configurations, finite paths and runs in $\pCFG$, respectively, 
dropping the index $\pCFG$ when known from the context.

@@

%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.5, inner sep = 1pt] {$\frac{3}{4}$} (prob|-dum1) -- node[auto] 
%{x:=x-1}
%(while|-dum1)--(while);
%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.5, inner sep = 1pt] {$\frac{1}{4}$} (prob|-dum2) 
%-- node[auto,swap] {x:=x+1} (while|-dum2)--(while);
%\draw[tran] (dif) --  (demon-|while) -- node[auto, pos=0.2] {x:=x+1} (while);
%\draw[tran] (aif) --  (angel-|while) -- node[auto, swap, pos=0.2] {x:=x+1} 
%(while);
%\draw[tran] (delse) -- node[auto] {x:=x-1} (delse-|while.305) --  (while.-55);
%\draw[tran] (aelse) -- node[auto, swap] {x:=x-1} (aelse-|while.55) --  
%(while.55);
\end{tikzpicture}
%\caption{}
\end{subfigure}
\caption{An example \APP{} and its associated 
pCFG. Probabilistic locations are depicted by circles, with probabilities given 
on outgoing 
transitions, demonic locations are denoted by diamonds. Transitions are 
labelled by their effects. Location $\loc_0$ is 
initial and $\loc_7$ is terminal.}
\label{fig:invariant-running}
\end{figure}

>>>>

\vspace{-1em}
\subsection{Basic Notions}
\vspace{-0.5em}
For a set $A$ we denote by $|A|$ the cardinality of $A$. We denote by $\Nset$,
$\Nset_0$, $\Zset$, and $\Rset$ the sets of all positive integers, non-negative
integers, integers, and real numbers, respectively. We assume basic knowledge 
of matrix calculus.
We use boldface notation for
vectors, e.g. $\vec{x}$, $\vec{y}$, etc., and we denote an $i$-th component of a
vector $\vec{x}$ by $\vec{x}[i]$. 
%For the purpose of matrix calculations we 
%assume that (non-transposed) vectors are row vectors. If $\vec{v},\vec{v}'$ 
%are 
%$n$ and $m$ dimensional vectors, respectively, then $(\vec{v},\vec{v}')$ is an 
%$(n+m)$-dimensional vector obtained by ``concatenation'' of $\vec{v}$ and 
%$\vec{v}'$. 
We identify 1-dimensional vectors with numbers. For an 
$n$-dimensional vector 
$\vec{x}$, index $1 \leq i\leq n$, and number $a$ we denote by $\vec{x}(i\leftarrow a)$ 
a 
vector $\vec{y}$ such that $\vec{y}[i]=a$ and $\vec{y}[j]=\vec{x}[j]$ for all 
$1\leq j \leq n$, $j\neq i$.
For comparison of vectors (e.g. as in $\vec{x}\leq \vec{y}$), we consider componentwise 
comparison. 
For comparing functions $f,g$ with the same domains, we write $f\leq g$ 
if $f(x)\leq g(x)$ for all $x$ in the domain. 
%\smallskip\noindent{\em Variables.}
Throughout the paper we fix a countable set of variables $\vars$. 
We consider some arbitrary but fixed linear order on the set of all variables, 
hence we write 
$\vars=\{x_1,x_2,x_3,\dots\}$. 
%A \emph{valuation} is a partial function 
%$\val\colon 
%\vars \dashrightarrow \Rset$ assigning a unique number to some of the 
%variables. A \emph{domain} of a valuation $\val$, denoted by %$\dom(\val)$, is 
%the set of all variables $x$ such that $\val(x)$ is defined. Given a %set 
%$\pvars\subseteq \vars$ of variables, we denote by $\vals(\pvars)$ the %set of 
%all valuations $\val$ with $\dom(\val)=\pvars$.

@@

\smallskip\noindent{\em Expressions.}
An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
is an expression in the standard programming-language sense, i.e. a formula 
built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
numerical operators from some fixed finite set. Each expression $E$ over 
$\{x_1,\dots,x_n\}$ 
determines a function which for each $m$-dimensional vector $\vec{x}$, where 
$m\geq n$,  
returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
Slightly abusing our notation, we denote this function also by $E$ and the 
value of this function on argument $\vec{x}$ by $E(\vec{x})$. We do not a priori fix a concrete set of operators that can be used to form expressions. However, in order to ensure that semantics of probabilistic programs with real-valued variables is defined correctly, we impose the following two conditions on the set of expressions used in each program:
%\begin{enumerate}
(1)
For each expression $E$ over variables $\{x_1,\dots,x_n\}$ and each $n$-dimensional vector $\vec{x}$ the value $E(\vec{x})$ is well defined.\footnote{Our results can be easily extended to programs where encountering an expression of undefined value, such as division by zero, triggers an exception which terminates the program, but we abstract away from such details for the sake of clarity.}
(2)
The function defined by each expression $E$ is Borel-measurable (for definition of Borel-measura\-bility, see, e.g.~\cite{Billingsley:book}).
%\end{enumerate}
%
From measure theory it is known that these conditions hold in particular for programs where expressions are build using the standard arithmetic operators of addition, subtraction, multiplication, and division (provided that expressions evaluating to zero are not allowed as divisors).

@@

\noindent{\em Probabilistic Control Flow Graphs.}
%There are several ways in which one can express the semantics of \APP's with
%(angelic and demonic) 
%non-determinism
We consider standard operational semantics of \PP{}s defined via an 
uncountable state-space
Markov decision process (MDP) (uncountable due to real-valued variables).
That is, we associate to each program 
a certain stochastic process.
To define this process, we first define so called 
\emph{probabilistic control flow graphs}~\cite{CFG16}.

\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinitset,\transitions,\updates,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into four 
pairwise
disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
non-deterministic branching, 
probabilistic branching, deterministic, and assignment locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinitset$ is a set of initial 
\emph{assignment vectors};
% POKRACOVAT ODSUD
\item $\transitions\subseteq\locs\times\locs$ is a transition relation;
%whose members are tuples of
%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
%variable index}, and $u$ 
%is an 
%\emph{update element}, which can 
%be one of the following mathematical objects: %\begin{itemize}
%%\item 
%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%%\item 
%(b)~a distribution $d\in \mathcal{D}$; or
%%%\item 
%(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\updates$ is a function assigning to each transition outgoing from an 
assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
\emph{target 
	variable index} and $\up$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~a Borel-measurable function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%\item 
(c)~a set $R\subseteq \Rset$ (representing a non-deterministic update).
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$;
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

We assume that each location has at least one outgoing transition.
Also, for every deterministic location $\ell$ we assume the following: if
$\tau_1,\dots,\tau_k$ are all transitions outgoing from $\ell$, then $G(\tau_1)
\vee \dots \vee G(\tau_k) \equiv \mathit{true}$ and $G(\tau_i) \wedge G(\tau_j)
\equiv \mathit{false}$ for each $1\leq i < j \leq k$. For each 
distribution $d$ appearing in the 
pCFG we assume the following features are known: expected value $\expv[d]$ of 
$d$ and a set $SP_d$ containing the \emph{support} of $d$. The support is the  
smallest 
closed set of real numbers whose complement has probability zero 
under $d$.\footnote{In particular, a support of a \emph{discrete} probability 
	distribution $d$ is simply the at most countable set of all points on a 
	real 
	line that have positive probability under $d$.} Finally, we assume that 
	each 
assignment location has at most (and thus exactly) one outgoing transition. The 
translation from probabilistic programs to the corresponding pCFG is 
standard~\cite{CFNH16:prob-termination}, and the details are presented 
in~\AppendixMaterial.
%Finally, we assume that for each 
%transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
%deterministic. This is just a technical assumption yielding no loss of 
%generality, and it somewhat simplifies notation.
%And we assume that each coordinate of $\mathcal{D}$ represents an integrable
%random variable (i.e., the expected value of the absolute value of the random
%variable exists).
\end{definition}
\noindent{\em Configurations.}
A \emph{configuration} of a pCFG $\pCFG$ is a tuple $(\ell,\vec{x})$,
where $\ell$ is a location of $\pCFG$ and $\vec{x}$ is an 
$|\pvars|$-dimensional vector.
We say that a transition $\tau$ is \emph{enabled} in a configuration
$(\ell,\vec{x})$ if $\ell$ is the source location of $\tau$ and in addition,
${\vec{x}}\models G(\tau)$ provided that $\ell$ is deterministic. 
%A 
%configuration 
%$(\loc,\vec{x})$ is non-deterministic/probabilistic/deterministic/assignment 
%one if 
%$\loc$ is 
%non-deterministic/probabilistic/deterministic/assignment one, respectively.

\noindent{\em Executions and reachable configurations.}
We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
configuration $(\loc,\vec{x})$ if 
%A \emph{finite path} (or
%\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
%configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
%for each
%$0 \leq i < k$ 
there is a transition  $\tau=(\loc,\loc')$ enabled in
$(\loc,\vec{x})$ 
%such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
%satisfies one of the following:
and $\vec{x}'$ satisfies the following:
\begin{compactitem}
\item if $\loc$ is not an assignment location, then $\vec{x}'=\vec{x}$;
\item if $\loc$ is an assignment location with $\updates(\tau)=(j,\up)$, then 
$\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ satisfies one of the 
following 
depending on the type of $\up$:
%\begin{compactitem}
%\item 
(a)~if $\up$ is a Borel-measurable function, then $a=\up(\vec{x})$;
%\item 
(b)~if $\up$ is an integrable\footnote{A distribution on some numerical 
domain 
is integrable if its expected value exists and is finite. In particular, each 
Dirac distribution is integrable.} distribution $d$, then 
$a\in \support(d)$; and
%\item
(c)~if $\up$ is a set, then $a$ is some element of $\up$.
%\end{compactitem}
\end{compactitem}
A \emph{finite path} (or
\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
$\loc_0=\locinit$, $\vec{x}_0\in\vecinitset$, and 
%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
for each
$0 \leq i < k$ the configuration $(\loc_{i+1},\vec{x}_{i+1})$ is a successor of 
$(\loc_i,\vec{x}_i)$.
A \emph{run} (or \emph{execution}) in
$\pCFG$ is an infinite sequence of configurations whose every finite
prefix is a finite path.
A configuration $(\loc,\vec{x})$ is {\em reachable} from the initial 
configuration
$(\locinit,\vecinit)$ (where, $\vecinit\in \vecinitset$)
if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
$(\loc,\vec{x})$. We denote by $\Conf_\pCFG,\Fpath_\pCFG$ and $\Run_\pCFG$ the 
sets of all configurations, finite paths and runs in $\pCFG$, respectively, 
dropping the index $\pCFG$ when known from the context.

@@

%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.5, inner sep = 1pt] {$\frac{3}{4}$} (prob|-dum1) -- node[auto] 
%{x:=x-1}
%(while|-dum1)--(while);
%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.5, inner sep = 1pt] {$\frac{1}{4}$} (prob|-dum2) 
%-- node[auto,swap] {x:=x+1} (while|-dum2)--(while);
%\draw[tran] (dif) --  (demon-|while) -- node[auto, pos=0.2] {x:=x+1} (while);
%\draw[tran] (aif) --  (angel-|while) -- node[auto, swap, pos=0.2] {x:=x+1} 
%(while);
%\draw[tran] (delse) -- node[auto] {x:=x-1} (delse-|while.305) --  (while.-55);
%\draw[tran] (aelse) -- node[auto, swap] {x:=x-1} (aelse-|while.55) --  
%(while.55);
\end{tikzpicture}
%\caption{}
\end{subfigure}
%\vspace{-3em}
\caption{An example \APP{} and its associated 
pCFG. Probabilistic locations are depicted by circles, with probabilities given 
on outgoing 
transitions, demonic locations are denoted by diamonds. Transitions are 
labelled by their effects. Location $\loc_0$ is 
initial and $\loc_7$ is terminal.}
\label{fig:invariant-running}
\end{figure}
